---
title: "Cross-validation"
author: "Mauricio Garcia-Tec, Natalia Zuñiga-Garcia"
date: "`r Sys.Date()`"
output:
  html_document: default
  html_vignette: default

vignette: >
  %\VignetteIndexEntry{Cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


We show a simple framework to perform cross-validation and optain optimal parameters for the density estimation. 

## The data

We will use one of the datasets that we have included with package, which includes the counts of rides between 9:00pm-10:00pm during the Saturdays of 5 consecutive weeks. We will do 5-fold cross-validation to tune the hyperparameters of our model.

```{r, message = FALSE, warning = FALSE}
library(knitr)
library(pldensity)
library(dplyr)
library(progress)
```

```{r}
data("austin5w") 
austin5w <- austin5w %>% 
  select(start_location_long, start_location_lat, started_on, week) 
summary(austin5w)
```

```{r}
austin5w %>% 
  head() %>%
  kable()
```

```{r}
austin5w %>% 
  tail() %>%
  kable()
```


## Density on the entire set

Define data as a matrix

```{r}
x <- austin5w %>% 
  select(start_location_long, start_location_lat) %>% 
  data.matrix()
```

```{r, cache = TRUE}
mod_all <- dpn_init(
  nparticles = 250,
  alpha = 25,
  lambda = colMeans(x),
  kappa = .01,
  nu = 2,
  Omega =  0.01 ^ 2 * diag(2)
) 
```

Train the model

```{r, cache=TRUE}
mod_all <- mod_all %>% 
  dpn_mix(x)
```

Visualize the result 

```{r, cache=TRUE, fig.dim=c(8,8)}
spatial_plot(mod_all, newx = x, polygons = TRUE, resol = 50, contour_levels = 500)
```


## Parameters for cross-validation

```{r}
params <- expand.grid(
  alpha = c(25, 50, 100, 200),
  omega_scale = c(.0005, .001, .01, .1),
  kappa = c(.0005, .001, .01, .1),
  nparticles = c(100, 250, 500)
)
```

```{r}
params %>% 
  head() %>% 
  kable()
```


## Template for cross-validation

This is a template for doing cross-validation, although we do not evaluate it for this demo.

```{r, eval = FALSE}
lambda <- c(mean(austin5w$start_location_long), mean(austin5w$start_location_lat))
params$test_ll_mean <- 0
params$test_ll_sd <- 0
pb <- progress_bar$new(total = 5 * nrow(params)) # progress bar
pb$tick()
for (i in 1:nrow(params)) {
  mod <- dpn_init(
    nparticles = params$nparticles[i],
    alpha = params$alpha[i],
    lambda = lambda,
    kappa = params$kappa[i],
    nu = 2,
    Omega =  params$omega_scale[i] ^ 2 * diag(2)
  )
  test_ll <- numeric(5)
  for (w in 1:5) {
    train <- austin5w %>% 
      filter(week == w) %>% 
      select(start_location_long, start_location_lat) %>% 
      data.matrix()
    test <- austin5w %>% 
      filter(week != w) %>% 
      select(start_location_long, start_location_lat) %>% 
      data.matrix()
    mod_trained <- mod %>% 
      dpn_mix(train)
    test_ll[w] <- mod_trained %>% 
      dpn_eval(test, nparticles = params$nparticles[i]) %>% 
      mean()
    pb$tick() # update progress bar
  }
  params$test_ll_mean[i] <- mean(test_ll)
  params$test_ll_sd[i] <- sd(test_ll)
}
saveRDS(params, "../data/cv_results.RDS")
```





