---
title: "Fit dynamic distributions"
author: "Mauricio Tec & Natalia Zuniga"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-7}
---

We'll see an example of time-varying mixture of normal distributions and how to select the best model.

```{r}
set.seed(9999)
library(plot3D)
library(pldensity)
library(mvtnorm)
library(ggplot2)
```

```{r}
time_periods <- 15
mu1_mean <- c(
  seq(-2.5, -1.3, length.out = time_periods %/% 3), 
  rep(-1.5, time_periods %/% 3),
  seq(-1.5, -2, length.out = time_periods %/% 3))
mu2_mean <- c(
  seq(2.5, 1.3, length.out = time_periods %/% 3), 
  rep(1.5, time_periods %/% 3),
  seq(1.5, 2, length.out = time_periods %/% 3))
```

```{r}
simdata <- vector("list", time_periods)
for (t in 1:time_periods) {
  source1 <- rnorm(50, mu1_mean[t])
  source2 <- rnorm(50, mu2_mean[t])
  simdata[[t]] <- matrix(c(source1, source2)[sample(100)], ncol = 1) # aleatorise order, matrix needed
}
```

This is the real density

```{r}
resol <- 100
x <- seq(-7, 7, length.out = 100)
y <- 1:time_periods
z <- matrix(0, resol, time_periods)
for (t in 1:time_periods) {
  for (i in 1:resol) {
    z[i, t] <- 0.5 * dnorm(x[i], mu1_mean[t]) + 0.5 * dnorm(x[i], mu2_mean[t])
  }
}
```


```{r, fig.width=7, fig.height=7}
persp3D(x, y, z, theta = 20, phi = 45, 
      border = 1, colkey = FALSE,  zlim = c(0, .4),
      xlab = "x", ylab = "time", zlab = "density",
      ticktype = "detailed",
      main = "true density")
```



### These are independent fits

```{r}
set.seed(110104)
nparticles = 250
mod0 <- ddpn_init(
  nparticles = nparticles, 
  lambda = 0, 
  kappa = 0.01,
  nu = 3,
  Omega = matrix(5),
  alpha = 3)
```

```{r}
mods_ind <- vector("list", time_periods)
for (t in 1:time_periods) {
  mods_ind[[t]] <- ddpn_mix(mod0, simdata[[t]])
}
```

Let's plot the results

```{r}
x <- seq(-7, 7, length.out = 100)
y <- 1:time_periods
z_ind <- matrix(0, resol, time_periods)
for (t in 1:time_periods) {
  z_ind[ ,t] <- ddpn_eval(mods_ind[[t]]$updated_model, matrix(x, ncol = 1), nparticles = nparticles) 
}
```

We see that after time period one the model is not captured very well.

```{r, fig.width=7, fig.height=7}
ribbon3D(x, y, z_ind, 
      theta = 20, phi = 45, border = 1, zlim = c(0, .4),
      xlab = "x", ylab = "time", zlab = "density",
      ticktype = "detailed", colkey = FALSE,
      main = "independent fit")
```

### These are dynamic (correlated) fits


```{r}
set.seed(110104)
nparticles = 250
mod0 <- ddpn_init(
  nparticles = nparticles, 
  lambda = 0, 
  kappa = 0.01,
  nu = 3,
  Omega = matrix(5),
  alpha = 3,
  rho = 0.25)
```

```{r}
mods_cor <- vector("list", time_periods) 
mods_cor[[1]] <- ddpn_mix(mod0, simdata[[1]])
for (t in 2:time_periods) {
  mods_cor[[t]] <- ddpn_mix(mods_cor[[t - 1]]$updated_model, simdata[[t]])
}
```


Let's plot the results

```{r}
x <- seq(-7, 7, length.out = 100)
y <- 1:time_periods
z_cor <- matrix(0, resol, time_periods)
for (t in 1:time_periods) {
  z_cor[ ,t] <- ddpn_eval(mods_cor[[t]]$updated_model, matrix(x, ncol = 1), nparticles = nparticles) 
}
```

We see that after time period one the model is not captured very well.

```{r, fig.width=7, fig.height=7}
ribbon3D(x, y, z_cor, 
      theta = 20, phi = 45, border = 1, zlim = c(0, .4),
      xlab = "x", ylab = "time", zlab = "density",
      ticktype = "detailed", colkey = FALSE,
      main = "correlated fit")
```

### This how we do model selection

We can compare the quality with the Bayes factor. We can compare different values of $\rho$ with the independent fit model.


```{r}
nparticles = 250
logmarginal <- data.frame()
for (r in seq(0.1, 0.9, .1)) {
  set.seed(110104)
  m_llhood <- numeric(time_periods)
  mod0 <- ddpn_init(
    nparticles = nparticles, 
    lambda = 0, 
    kappa = 0.01,
    nu = 3,
    Omega = matrix(5),
    alpha = 3,
    rho = r)
  mods_cor <- vector("list", time_periods) 
  mods_cor[[1]] <- ddpn_mix(mod0, simdata[[1]])
  m_llhood[1] <- mods_cor[[1]]$marginal_loglikelihood
  for (t in 2:time_periods) {
    mods_cor[[t]] <- ddpn_mix(mods_cor[[t - 1]]$updated_model, simdata[[t]])
    m_llhood[t] <- mods_cor[[t]]$marginal_loglikelihood
  }
  logmarginal <- rbind(logmarginal, data.frame(period = 1:time_periods, rho = r, logmarginal = m_llhood))
}
```

```{r}
ind_logmarginal <- numeric(time_periods)
for (t in 1:time_periods) {
  ind_logmarginal[t] <- mods_ind[[t]]$marginal_loglikelihood 
}
logmarginal$rho <- ordered(logmarginal$rho)
logmarginal$logbayes_factor <- logmarginal$logmarginal - ind_logmarginal
```


```{r, fig.width=6, fig.height=3}
ggplot(logmarginal, aes(x = period, y = logbayes_factor, colour = rho)) +
  geom_line() +
  labs(title = "Bayes factor vs Independent fit") +
  geom_hline(yintercept = 0)
```

